Symbolic On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-01 main.tex
check point

@@

\section{Preliminary}

@@

\subsection{Guarded Kleene Algebra With Tests}

\subsection{GKAT Coalgebra}

% definition
GKAT coalgebra~\cite{smolka_GuardedKleeneAlgebra_2020,schmid_GuardedKleeneAlgebra_2021}, is a coalgebraic systems for GKAT.
Specifically GKAT coalgebras over a alphabet \(K, B\) are coalgebras over the following functor:
\[G(S) ≜ (2 + S × K)^{\At_B},\] 
where \(2 ≜ \{\accept, \reject\}\). Intuitively given a state \(s ∈ S\) and an atom \(α ∈ \At\), \(δ(s, α)\) will deterministically execute one of the following: reject \(α\), denoted as \(δ(s, α) = \reject\); accept \(α\), denoted \(δ(s, α) = \accept\); or transition to a state \(s' ∈ S\) and execute action \(p ∈ K\), denoted as \(δ(s, α) = (s', p)\).

This deterministic behavior contrast that of Kleene coalgebra with tests~\cite{kozen_CoalgebraicTheoryKleene_2017}, where for each atom, the state can accept or  reject the atom (but not both), yet the state can also non-deterministically transition to multiple different state via the same atom, while executing different actions.
% TODO: I think this belongs in the intro
As we will see later, the deterministic behavior of GKAT coalgebra not only enables a further optimized symbolic algorithm than KCT~\cite{pous_SymbolicAlgorithmsLanguage_2015}, but also present challenges. Specifically, GKAT coalgebra requires normalization to compute finite trace equivalences~\cite{smolka_GuardedKleeneAlgebra_2020}, where we will remove all the state that cannot lead to acceptance. 
It seems like we need to traverse the entire automaton to identify these ``dead states'', however we have shown that these dead state detection can be invoked lazily, only when discrepancy between bisimulation are found.

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}. However, recall that principle sub-coalgebra \(⟨s⟩_S\) models reachable states of \(s\) in coalgebra \(S\). Thus the classical definition is equivalent to the following:
\begin{definition}[liveness of states]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists a \(α ∈ \At\) s.t. \(δ(s, α) = \accept\). A state \(s'\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\). A state \(s'\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{definition}
This alternative liveness definition can help us formally prove important theorems regarding reachability and liveness without performing induction on traces. We can show the following lemmas as examples:
\begin{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{lemma}
\begin{proof}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) by definition. 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(⟨s'⟩\) is also dead.
\end{proof}

\begin{theorem}[homomorphism perserves liveness]\label{thm:hom-preserve-liveness}
    Given a homomorphism \(h: S → T\) and a state \(s ∈ S\):
    \[\text{\(s\) is live} ⟺ \text{\(h(s)\) is live}\]
\end{theorem}

\begin{proof}
    Because homomorphic image preserves principle sub-GKAT coalgebra (\cref{thm:homo-img-preserve-principle-sub-coalg})
    \[h(⟨s⟩_S) = ⟨h(s)⟩_T;\]
    therefore for any state \(s' ∈ S\):
    \[s' ∈ ⟨s⟩_S ⟺ h(s') ∈ h(⟨s⟩_S) ⟺ h(s') ∈ ⟨h(s)⟩_T.\]
    And because \(s'\) is accepting if and only if \(h(s')\) accepting by definition of homomorphism; then \(⟨s⟩_S\) contains an accepting state if and only if \(h(⟨s⟩_S) = ⟨h(s)⟩_T\) contains an accepting state. 
    Therefore \(s\) is live in \(S\) if and only if \(h(s)\) is live in \(T\).
\end{proof}

\begin{corollary}[sub-coalgebra perserves liveness]\label{thm:sub-coalg-preserve-liveness}
    Given a sub-coalgebra \(S' ⊑ S\), then for all states \(s ∈ S'\),
    \[\text{\(s\) is live in \(S'\)} ⟺ \text{\(s\) is live in \(S\)}.\]
\end{corollary}

\begin{proof}
    take the homomorphism \(h\) in \cref{thm:hom-preserve-liveness} to be the inclusion homomorphism \(i: S' → S\).
\end{proof}

\begin{corollary}[bisimulation preserves liveness]\label{thm:bisim-preserve-liveness}
    If there exists a bisimulation \(∼\) between GKAT coalgebra \(S\) and \(T\) s.t. \(s ∼ t\) for some states \(s ∈ S\) and \(t ∈ T\), then \(s\) and \(t\) has to be either both accepting, both live or both dead.
\end{corollary}

\begin{proof}
    Because for a \(∼\) is a bisimulation when both \(π₁: {∼} → S\) and \(π₂: {∼} → T\) are homomorphisms.
    Therefore, 
    \begin{align*}
        s \text{ is live in } S 
        & ⟺ π₁((s, t)) \text{ is live in } S 
            & π₁((s, t)) = s\\  
        & ⟺ (s, t) \text{ is live in } {∼} 
            & \text{\(π₁\) is a homomorphism}\\  
        & ⟺ π₂((s, t)) \text{ is live in } T 
            & \text{\(π₂\) is a homomorphism} \\
        & ⟺ t \text{ is live in } T 
            & π₂((s, t)) = t
    \end{align*}
\end{proof}

% The signature of GKAT coalgebra \(G\) is a simple polynomial functor~\cite[Definition 2.2.1]{jacobs_IntroductionCoalgebraMathematics_2016}, implying that it preserves the inclusion function:
% \begin{lemma}%used in the comments below
%     Given a inclusion \(i: S' → S\) for \(S' ⊆ S\), then \(G(i): G(S') → G(S)\) is also a inclusion in the sense that:
%     \[∀ m ∈ G(S'), G(i)(m) = m.\]
% \end{lemma}
% This implies that for a sub-GKAT coalgebra \(S' ⊑ S\), then the restriction of transition function will satisfy the the diagram for sub-coalgebra.
% \[
%     \begin{tikzcd}
%         S' \ar{d}{δ_S} \ar[hook]{r}{i} & S \ar{d}{δ_S} \\  
%         G(S') \ar[hook]{r}{G(i)} & G(S)
%     \end{tikzcd}
% \]
% Then by the uniqueness of transition for sub-coalgebra~\cite[Proposition 6.1]{rutten_UniversalCoalgebraTheory_2000}, the transition of \(S'\) is necessarily the restriction of \(δ_S\). A nice consequence of this observation is that the ordering between sub-GKAT coalgebra is exactly the subset ordering.
% \begin{lemma}[ordering in sub-GKAT coalgebra]\label{thm:sub-GKAT coalgebra-is-subset}
%     Given two sub-GKAT coalgebra \(S₁ ⊑ S\) and \(S₂ ⊑ S\), then 
%     \[S₁ ⊑ S₂ ⟺ S₁ ⊆ S₂.\]
% \end{lemma}

% \begin{proof}
%     \(⟹\) direction is true by definition, and \(⟸\) direction is true because both \(S₁\) and \(S₂\) are sub-GKAT coalgebra, thus their transition functions are both restrictions of \(δ_S\). Hence the commutativity of the following diagram can be verified by computation:
%     \[
%     \begin{tikzcd}
%         S₁ \ar[hook]{r}{i} \ar{d}{δ_S} & S₂ \ar[hook]{r}{i} \ar{d}{δ_S} & S \ar{d}{δ} \\  
%         G(S₁) \ar[hook]{r}{G(i)} & G(S₂) \ar[hook]{r}{G(i)} & G(S)
%     \end{tikzcd}
% \]
% \end{proof}

% trace semantics
\subsection{Normalization And Trace Semantics}

(Possibly infinite) trace model \(𝒢_ω\) is the natural semantics of states in a GKAT coalgebra, specifically it forms the final coalgebra of GKAT coalgebras~\cite{schmid_GuardedKleeneAlgebra_2021}.
The finality of the model means that every state in any GKAT coalgebra \(S\) can be assigned a semantics under the unique homomorphism \(⟦-⟧^{ω}_{S}: S → 𝒢_ω\); and such semantical equivalences can indeed be identified by bisimulation~\cite{schmid_GuardedKleeneAlgebra_2021}:
\[⟦s⟧^{ω}_{S} = ⟦t⟧^{ω}_{T} ⟺ \text{exists a bisimulation \({∼} ⊆ S × T\), s.t. \(s ∼ t\)}.\]

The infinite trace equivalences is relatively easy to compute efficiently, as bisimulation is, in general, compatible with derivative based computation on-the-fly algorithm~\cite{kozen_CoalgebraicTheoryKleene_2017,almeida_DecidingKATHoare_2012,pous_SymbolicAlgorithmsLanguage_2015}. 
However, the \emph{finite} trace model \(𝒢\) is only the final coalgebra of GKAT coalgebras without dead states, which we call \emph{normal GKAT coalgebra}~\cite{smolka_GuardedKleeneAlgebra_2020}. Fortunately every GKAT coalgebra can be normalized by rerouting all the transition from dead states to rejection
\begin{align*}
    \norm(δ_S) & : S → G(S) \\
    \norm(δ_S)(s, α) & ≜ \begin{cases}
        \reject & \text{if } δ_S(s, α) = (s', p) \text{ and \(s'\) is dead}\\
        δ_S(s, α) & \text{otherwise}
    \end{cases}
\end{align*}
We denote the normalized coalgebra \((S, \norm(δ_S))\) as \(\norm(S)\).
This means that the finite trace semantics \(⟦-⟧\) is the unique coalgebra homomorphism \(\norm(S) → \norm(𝒢)\). Hence the finite trace equivalence between \(s ∈ S\) and \(t ∈ T\) can be computed by first normalizing \(S\) and \(T\), then decide whether there is a bisimulation on \(\norm(S)\) and \(\norm(T)\) that includes \((s, t)\).
For a more intuitive account for the trace semantics, we refer the reader to the work of \Citeauthor{smolka_GuardedKleeneAlgebra_2020}~\cite{smolka_GuardedKleeneAlgebra_2020}.

Then by the finality of GKAT colagebra, the soundness and completeness of bisimulation is a simple corollary in universal coalgebra~\cite{smolka_GuardedKleeneAlgebra_2020,jacobs_IntroductionCoalgebraMathematics_2016,rutten_UniversalCoalgebraTheory_2000}. No matter through explicit construction~\cite{smolka_GuardedKleeneAlgebra_2020} or (weak) pullback~\cite{jacobs_IntroductionCoalgebraMathematics_2016,rutten_UniversalCoalgebraTheory_2000}, both completeness proof shows that the language equivalence: \[{≡} ≜ \{(s,t) ∣ ⟦s⟧_S = ⟦t⟧_T\}\] is indeed a bisimulation. 
This result, together with soundness result \(s ∼ t ⟹ ⟦s⟧_S = ⟦t⟧_T\) means that the language equivalence \(≡\) is indeed the largest bisimulation, which allows us to work with bisimulation equivalence instead of bisimulation.

>>>>

% page number
\pagestyle{plain}

@@

\section{Background on Coalgebra and GKAT}

@@

\subsection{GKAT and Its Coalgebra}

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\Bool\) and expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \Bool 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \Bool ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\); \(e;f\) is the sequencing of expression \(e\) and \(f\); finally \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with tests.

However, the semantics of GKAT can also be obtained coalgebraically, specifically, with GKAT coalgebra~\cite{smolka_GuardedKleeneAlgebra_2020,schmid_GuardedKleeneAlgebra_2021}.
Formally, GKAT coalgebras over primitive actions \(K\) and primitive tests \(T\) are coalgebras over the following functor:
\[G(S) ≜ (2 + S × K)^{\At_B},\] 
where \(2 ≜ \{\accept, \reject\}\). Intuitively given a state \(s ∈ S\) and an atom \(α ∈ \At\), \(δ(s, α)\) will deterministically execute one of the following: reject \(α\), denoted as \(δ(s, α) = \reject\); accept \(α\), denoted \(δ(s, α) = \accept\); or transition to a state \(s' ∈ S\) and execute action \(p ∈ K\), denoted as \(δ(s, α) = (s', p)\).

This deterministic behavior contrast that of Kleene coalgebra with tests~\cite{kozen_CoalgebraicTheoryKleene_2017}, where for each atom, the state can accept or  reject the atom (but not both), yet the state can also non-deterministically transition to multiple different state via the same atom, while executing different actions.
% TODO: I think this belongs in the intro
As we will see later, the deterministic behavior of GKAT coalgebra not only enables a more versatile symbolic algorithm than KCT~\cite{pous_SymbolicAlgorithmsLanguage_2015}, but also present challenges. 
Specifically, GKAT coalgebra requires normalization to compute finite trace equivalences~\cite{smolka_GuardedKleeneAlgebra_2020}, where we will remove all the state that cannot lead to acceptance, which are called ``dead states''.

In previous works, these dead states are detected after all the necessary state and transition is computed.
This approach requires storing all the information of the coalgebra in memory, which do not allow on-the-fly computation, which can not only terminate whenever a counter-example is found, but can also erase past states from memory.

However, to truly understand the on-the-fly algorithm, we will first need to define ``dead states'', and its role in defining the coalgebraic semantic of GKAT.

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}. However, it is straightforward to show that principle sub-coalgebra \(⟨s⟩_S\) exactly correspondents to reachable states of \(s\) in coalgebra \(S\). 
Thus, the classical definition is equivalent to the following:
\begin{definition}[liveness of states]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists a \(α ∈ \At\) s.t. \(δ(s, α) = \accept\). A state \(s'\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\). A state \(s'\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{definition}
This alternative liveness definition can help us formally prove important theorems regarding reachability and liveness without performing induction on traces. We can show the following lemmas as examples:
\begin{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{lemma}
\begin{proof}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) by definition. 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(⟨s'⟩\) is also dead.
\end{proof}

\begin{theorem}[homomorphism perserves liveness]\label{thm:hom-preserve-liveness}
    Given a homomorphism \(h: S → U\) and a state \(s ∈ S\):
    \[\text{\(s\) is live} ⟺ \text{\(h(s)\) is live}\]
\end{theorem}

\begin{proof}
    Because homomorphic image preserves principle sub-GKAT coalgebra (\Cref{thm:homo-img-preserve-principle-sub-coalg})
    \[h(⟨s⟩_S) = ⟨h(s)⟩_U;\]
    therefore for any state \(s' ∈ S\):
    \[s' ∈ ⟨s⟩_S ⟺ h(s') ∈ h(⟨s⟩_S) ⟺ h(s') ∈ ⟨h(s)⟩_U.\]
    And because \(s'\) is accepting if and only if \(h(s')\) accepting by definition of homomorphism; then \(⟨s⟩_S\) contains an accepting state if and only if \(h(⟨s⟩_S) = ⟨h(s)⟩_U\) contains an accepting state. 
    Therefore, \(s\) is live in \(S\) if and only if \(h(s)\) is live in \(U\).
\end{proof}

The above theorem then leads to several interesting liveness preservation properties for important structures on coalgebras, like sub-coalgebra and bisimulation.

\begin{corollary}[sub-coalgebra perserves liveness]\label{thm:sub-coalg-preserve-liveness}
    For a sub-coalgebra \(S' ⊑ S\) and a state \(s ∈ S'\), \(s\) is live in \(S'\) if and only if \(s\) is live in \(S\).
\end{corollary}

\begin{proof}
    Let the homomorphism \(h\) in \cref{thm:hom-preserve-liveness} be the inclusion homomorphism \(i: S' → S\).
\end{proof}

\begin{corollary}[bisimulation preserves liveness]\label{thm:bisim-preserve-liveness}
    If there exists a bisimulation \(∼\) between GKAT coalgebra \(S\) and \(U\) s.t. \(s ∼ u\) for some states \(s ∈ S\) and \(u ∈ U\), then \(s\) and \(u\) has to be either both accepting, both live or both dead.
\end{corollary}

\begin{proof}
    Because for a \(∼\) is a bisimulation when both \(π₁: {∼} → S\) and \(π₂: {∼} → U\) are homomorphisms.
    Therefore, 
    \begin{align*}
        s \text{ is live in } S 
        & ⟺ π₁((s, u)) \text{ is live in } S \\        
        & ⟺ (s, u) \text{ is live in } {∼} \\  
        & ⟺ π₂((s, u)) \text{ is live in } U  \\
        & ⟺ u \text{ is live in } U. 
        \qedhere
    \end{align*}
\end{proof}

% trace semantics
\subsection{Normalization And Semantics}

(Possibly infinite) trace model \(𝒢_ω\) is the final coalgebra of GKAT coalgebras~\cite{schmid_GuardedKleeneAlgebra_2021}.
The finality of the model implies that every state in any GKAT coalgebra \(S\) can be assigned a semantics under the unique homomorphism \(⟦-⟧^{ω}_{S}: S → 𝒢_ω\); and such semantic equivalences can indeed be identified by bisimulation~\cite{schmid_GuardedKleeneAlgebra_2021}:
\(⟦s⟧^{ω}_{S} = ⟦t⟧^{ω}_{T}\) if and only if there exists a bisimulation \({∼} ⊆ S × T\), s.t. \(s ∼ t\).

The infinite trace equivalences can be directly computed with bisimulation on derivative, which supports on-the-fly algorithm as demonstrated by similar systems~\cite{kozen_CoalgebraicTheoryKleene_2017,almeida_DecidingKATHoare_2012,pous_SymbolicAlgorithmsLanguage_2015}. 
However, the \emph{finite} trace model \(𝒢\) is the final coalgebra of GKAT coalgebras without dead states, which we call \emph{normal GKAT coalgebra}~\cite{smolka_GuardedKleeneAlgebra_2020}. 
Fortunately every GKAT coalgebra can be normalized by rerouting all the transition from dead states to rejection.
Concretely, \(\norm(δ_S) : S → G(S)\) is defined as \(\norm(δ_S)(s, α) ≜ \reject\) when \(δ_S(s, α) = (s', p)\) and \(s'\) is dead in \(S\); and \(\norm(δ_S)(s, α) ≜ δ_S(s, α)\) otherwise.
We denote the normalized coalgebra \((S, \norm(δ_S))\) as \(\norm(S)\).

The finality of \(𝒢\) means that the finite trace semantics \(⟦-⟧\) is the unique coalgebra homomorphism \(\norm(S) → 𝒢\). 
Furthermore, the finite trace equivalence between \(s ∈ S\) and \(u ∈ U\) can be computed by first normalizing \(S\) and \(U\), then decide whether there is a bisimulation on \(\norm(S)\) and \(\norm(U)\) that includes \((s, t)\).
For more details on the finite trace semantics, we refer the reader to the work of \Citeauthor{smolka_GuardedKleeneAlgebra_2020}~\cite{smolka_GuardedKleeneAlgebra_2020}.

% Then by the finality of GKAT colagebra, the soundness and completeness of bisimulation is a simple corollary in universal coalgebra~\cite{smolka_GuardedKleeneAlgebra_2020,jacobs_IntroductionCoalgebraMathematics_2016,rutten_UniversalCoalgebraTheory_2000}. No matter through explicit construction~\cite{smolka_GuardedKleeneAlgebra_2020} or (weak) pullback~\cite{jacobs_IntroductionCoalgebraMathematics_2016,rutten_UniversalCoalgebraTheory_2000}, both completeness proof shows that the language equivalence: \[{≡} ≜ \{(s,t) ∣ ⟦s⟧_S = ⟦t⟧_T\}\] is indeed a bisimulation. 
% This result, together with soundness result \(s ∼ t ⟹ ⟦s⟧_S = ⟦t⟧_T\) means that the language equivalence \(≡\) is indeed the largest bisimulation, which allows us to work with bisimulation equivalence instead of bisimulation.
